---
title: "Project W2 Reproducible research"
author: "julianehf"
date: "29/06/2017"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## clear environment and set working directory

```{r}
rm(list=ls())
setwd('/Users/julianefarthouat/Documents/ETUDES/data science specialization/reproducible research/W2')
```

## Loading and preprocessing the data

Here we load and look at the data. 

```{r}
actdata = read.csv('activity.csv',header=T)
head(actdata)
str(actdata)
summary(actdata)
```


Let's transform dates into the R-date format and check if it did properly with the *str* function.

```{r}
actdata$date = as.Date(as.character(actdata$date),"%Y-%m-%d")
str(actdata)
```

### What is mean total number of steps taken per day?

1. We calculate the total number of steps taken per day.

```{r}
stepsperday = aggregate(steps~date,actdata,sum)
head(stepsperday)
```

2. We make an histogram of the total number of steps taken each day. 
```{r}
hist(stepsperday$steps,
     xlab='number of steps',breaks=50,main='Histogram of the total number of steps taken each day')
```

3. Calculate mean and median of total number of steps taken per day.
```{r}
meansteps=mean(stepsperday$steps)
mediansteps=median(stepsperday$steps)
```

Taken per day, the mean number of steps is `r meansteps` and the median number of steps is `r mediansteps`. 



## What is the average daily activity pattern?

1. We calculate the mean steps taken per interval averaged across all days. 

```{r}
stepsperint = aggregate(steps~interval,actdata,mean)
head(stepsperint)
```

Then we plot the time series of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).

```{r}
maxint = stepsperint[stepsperint$steps==max(stepsperint$steps),]$interval
max_x = with(stepsperint,which(maxint==stepsperint))

with(stepsperint, {
        plot(interval, steps,
                       type='b',ylab='mean steps per interval')
        text(maxint+800,steps[max_x],paste('maximum number of steps at interval ',maxint,sep=''))                
        })
```

2. The 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps is `r maxint`.

## Imputing missing values

1. Let's check how many missing values there are in the data. 
```{r}
sum(is.na(actdata$steps))
sum(is.na(actdata$date))
sum(is.na(actdata$interval))
```

There were `r sum(is.na(actdata$steps))` missing values of the *steps* variable. 

2. To fill the missing values, we replace it by the mean value of that 5-min interval. 

```{r}
actdata$steps2 = actdata$steps
for(i in c(1:nrow(actdata)  ) ){
        if(is.na(actdata[i,]$steps2)){
                int = actdata[i,]$interval
                tmp = stepsperint[stepsperint$interval==int,]$steps
                actdata[i,]$steps2=tmp
        }
}
```

Check that all missing values were remplaced: 
```{r}
sum(is.na(actdata$steps2))
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
actdataclean = actdata
actdataclean$steps=actdataclean$steps2
actdataclean$steps2 <- NULL
head(actdataclean)
```

4. Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. 
```{r}
stepsperdayclean = aggregate(steps~date,actdataclean,sum)
hist(stepsperdayclean$steps,
     xlab='number of steps',breaks=50,main='Histogram of the total number of steps taken each day')

```
```{r}
meansteps=mean(stepsperdayclean$steps)
meansteps=round(meansteps,0)
mediansteps=median(stepsperdayclean$steps)
```

Taken per day, the mean number of steps is `r meansteps` and the median number of steps is `r mediansteps`. These values do not differ from the estimates from the first part of the assignment. 

Let's observe the impact of imputing missing data on the estimates of the total daily number of steps.

```{r}
summary(stepsperday$steps)
summary(stepsperdayclean$steps)
```

The impact of imputing missing data on the estimates of the total daily number of steps seems very limited.

## Are there differences in activity patterns between weekdays and weekends?

1. To observe differences in activity patterns between weekdays and weekends, we create a new factor variable in the dataset with 2 levels: *weekday* and *weekend*.

```{r}
library(lubridate)
actdataclean$weekday = wday(actdata$date,label=T)
actdataclean$weekdayN = wday(actdata$date,label=F)
actdataclean$weekdayWE = actdataclean$weekdayN>6
actdataclean$weekdayWE[actdataclean$weekdayWE=='FALSE'] = 'weekday'
actdataclean$weekdayWE[actdataclean$weekdayWE=='TRUE'] = 'weekend'
actdataclean$weekdayWE = as.factor(actdataclean$weekdayWE)
head(actdataclean)
```

Now we plot the time series of the 5-minute interval and the average number of steps taken, averaged across all weekdays and all weekends. 

```{r}
datawe = aggregate(steps~interval*weekdayWE,actdataclean,mean)

par(mfrow=c(1,2))
with(datawe[datawe$weekdayWE=='weekday',], 
     plot(interval, steps,
          type='l',ylab='average number of steps taken per 5-minute interval',ylim=range(datawe$steps),main='WEEK DAYS') 
)
with(datawe[datawe$weekdayWE=='weekend',], 
     plot(interval, steps,
          type='l',ylab='average number of steps taken per 5-minute interval',ylim=range(datawe$steps),main='WEEK ENDS') 
)

```

And this is done, thank you for reading me ! :-) 
